---
title: "Effect of transmission type on MPG in cars"
author: "Marc T. Henry de Frahan"
output:
  pdf_document:
    fig_width: 3.5
    fig_height: 3
    number_sections: true
    fig_caption: true
---

# Executive summary

In this report, we analyze the relationship between transmission type
(automatic vs manual) and miles per gallon (outcome). Through a simple regression modeling looking just at transmission type, we find that: 

1. manual transmission is better for MPG;
2. manual transmission is about 7 MPG better and this difference is statistically significant (p-value < 0.05).

However, we also saw that there is a relationship between the
transmission type and the weight of the car. When confounding the
model with weight and transmission, we see that the transmission type
does not have a significant effect on the mpg.


# Exploratory data analysis

Let's first load the data and format the information we want from the
mtcars data set. We can plot the data using a violin plot to get an
idea of the differences between the transmission types. A first glance
at the data seems to indicate that the automatic transmission leads to
fewer miles per gallon. On the other hand, using a manual transmission
seems to lead to a better outcome though the spread is more. We will
perform linear regression to explore quantitatively the relationship
between transmission type and MPG.

```{r echo = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
df <- mtcars %>%
    mutate(transmission = ifelse(am==0,"automatic","manual")) %>%
    mutate(am = as.factor(am)) %>%
    mutate(transmission = as.factor(transmission))
```

# Regression models

## Simple linear regression model based on transmission
Let's perform the regression model to explore the relationship between
mpg and transmission type.
```{r echo = FALSE, message=FALSE, warning=FALSE}
fit <- lm(mpg ~ transmission, data = df)
b <- round(dfbetas(fit)[,2],3)
h <- round(hatvalues(fit),3)
```

### Model interpretation
Since we are looking at categorical data, the output coefficients
should be interpreted as intercepts. For automatic transmission the
intercept is `r fit$coef[1]`. For the manual transmission the
intercept is `r fit$coef[1]+fit$coef[2]` (the sum of the automatic
transmission intercept and the manual transmission coefficient). The
manual transmission coefficient indicates the change with respect to
the automatic transmission intercept. In this case it is positive,
which means that the intercept for manual transmission is greater than
the one for the automatic transmission. The difference is also
statistically significant since the p-value associated to this
coefficient ($`r summary(fit) $coeff[2,4]`$) is less than 0.05. This
implies that we can safely reject the null hypothesis, which is that
the means of the different transmission types are equal.


The implications of this model are that manual transmission is better
for MPG by about `r fit$coef[2]` MPG and this difference is
statistically significant.

### Residual and diagnostics

It is important to look at the residuals and other diagnostics to ensure that the model is ok. In the residual plot figure, we show that the residual does not follow any clear pattern, indicating that we are not missing a non-linearity in the model. In the residual diagnostics figures, we can see that the residuals closely follow a normal distribution. We can also see in the residual diagnostic figures that none of the data points have too much influence on the model. This can also be seen by looking at the dfbetas (max= $`r max(b)`$, min = $`r min(b)`$, and std = $`r sd(b)`$ ) and hatvalues (max= $`r max(h)`$, min = $`r min(h)`$, and std = $`r sd(h)`$ ) and noticing that none of the points are significantly different from each other. We conclude that the residual diagnostics do not invalidate our model.


## Confounding with the weight of the car

```{r echo = FALSE, message=FALSE, warning=FALSE}
fit.wt <- lm(wt ~ transmission, data = df)
```
Let's take into account one more variable, the weight of the car. We can see in the figure of weight versus transmission type that automatic cars tend to be heavier. A linear model of weight (outcome) on transmission type indicates that this is indeed the case. A manual car is predicted to be `r summary(fit.wt)$coeff[2,1]` lighter than a manual car. The difference is statistically significant (p-value = $`r summary(fit.wt)$coeff[2,4]`$ < 0.05). This implies that we should also look at the weight of the car as a predicting variable.

```{r echo = FALSE, message=FALSE, warning=FALSE}
fit.wt.am <- lm(mpg ~ transmission+wt, data = df)
```

This model confounding with the weight and transmission type indicates that there is a linear relationship of mpg on the weight (slope = $`r summary(fit.wt.am)$coeff[3,1]`$ ) and the transmission type does not have significant impact on the intercept (p-value = $`r summary(fit.wt.am)$coeff[2,4]`$ > 0.05).


# Conclusions 
For a simple linear model accounting only for transmission type , we can answer both our questions that explore the simple relationship between transmission type and mpg

1. The implications of our simple model are that manual transmission is better for MPG.
2. Manual transmission is about `r fit$coef[2]` MPG better and this difference is statistically significant (p-value = $`r summary(fit)$coeff[2,4]`$ < 0.05).

However, when confounding the model with weight and transmission, we
see that the transmission type does not have a significant effect on
the mpg.

\newpage

# Appendix
Here are the figures used in this document:

```{r echo = FALSE, fig.cap="MPG vs transmission type"}
ggplot(data=df, aes(x=transmission,y=mpg)) +
    geom_violin(aes(fill=transmission)) + 
    labs(x = "Transmission type",
         y="Miles per gallon (MPG)",
         title="Miles per gallon vs transmission");
```

```{r echo = FALSE, fig.cap="Residual plot"}
df <- df %>%
    mutate(residual = resid(fit))
ggplot(data=df, aes(x=transmission,y=residual)) +
    geom_violin(aes(fill=transmission)) + 
    labs(x = "Transmission type",
         y="Residual",
         title="Residual plot");
```

```{r echo = FALSE, fig.cap="Residual diagnostics"}
par(mfrow=c(1,1))
plot(fit)
```

```{r echo = FALSE, fig.cap="Weight vs transmission"}
ggplot(data=df, aes(x=transmission,y=wt)) +
    geom_violin(aes(fill=transmission)) + 
    labs(x = "Transmission type",
         y = "Weight",
         title="Weight vs transmission type");
```


The version history of this document can be found at the [GitHub repository page](https://github.com/marchdf/LinReg). Here is the full code used in this document.

```{r code=readLines(knitr::purl('./linreg_cars.Rmd', documentation = 1)), eval = FALSE}
```
