---
title: "Effect of Vitamin C dose and delivery method on guinea pig tooth growth"
author: "Marc T. Henry de Frahan"
output:
  pdf_document:
    fig_width: 3.5
    fig_height: 3
    number_sections: true
    fig_caption: true
---

# Executive summary

In this report, we analyze the relationship between tranmission type
(automatic vs manual) and miles per gallon (outcome). Through linear
regression models we find that: 

1. 
2. 

# Exploratory Data Analysis

Let's first load the data and format the information we want from the
mtcars data set. We can plot the data using a violin plot to get an
idea of the differences between the transmission types. A first glance
at the data seems to indicate that the automatic transmission leads to
fewer miles per gallon. On the other hand, using a manual transmission
seems to lead to a better outcome though the spread is more. We will
perform linear regression to explore quantitatively the relationship
between transmission type and MPG.

```{r echo = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
df <- mtcars[c("mpg","am")]
df <- df %>%
    mutate(transmission = ifelse(am==0,"automatic","manual")) %>%
    mutate(am = as.factor(am)) %>%
    mutate(transmission = as.factor(transmission))
str(df)
```

```{r echo = FALSE, fig.cap="MPG vs transmission"}
ggplot(data=df, aes(x=transmission,y=mpg)) +
    geom_violin(aes(fill=transmission)) + 
    labs(x = "Transmission type",
         y="Miles per gallon (MPG)",
         title="Miles per gallon vs transmission type");
```



# Regression models

Let's perform the regression model to explore the relationship between
mpg and transmission type.
```{r echo = FALSE, message=FALSE, warning=FALSE}
fit <- lm(mpg ~ transmission, data = df)
summary(fit)
```

Since we are looking at categorical data, the output coefficients
should be interpreted as intercepts. For automatic transmission the
intercept is `r fit$coef[1]`. For the manual transmission the
intercept is `r fit$coef[1]+fit$coef[2]` (the sum of the automatic
transmission intercept and the manual transmission coefficient). The
manual transmission coefficient indicates the change with respect to
the automatic transmission intercept. In this case it is positive,
which means that the intercept for manual transmission is greater than
the one for the automatic transmission. The difference is also
statistically significant since the p-value associated to this
coefficient (`r summary(fit)$coeff[2,4]`) is less than 0.05. This
implies that we can safely reject the null hypothesis, which is that
the means of the different transmission types are equal.



# Conclusions 
Under our assumptions, we can answer both our questions:

1. Increasing the dose of vitamin C increases the tooth growth.
2. The delivery method for vitamin C does not affect tooth growth.


\newpage

# Appendix
The version history of this document can be found at the [GitHub repository page](https://github.com/marchdf/LinReg). Here is the full code used in this document.

```{r code=readLines(knitr::purl('./linreg_cars.Rmd', documentation = 1)), eval = FALSE}
```
